\speculake: Hiding Malware in Speculative Execution

### 2018-05-08 paper/architecture.tex
target fn translated to speculative gadget

To verify the upper limit of speculative instructions, we instrumented our
trigger and payload programs to test a variable-length target function before it
communicated a signal to the real world via a cache side channel. If the cache
side channel revealed no signal in the real world, then we know the speculative
execution did not make it to the signal instructions before the mis-speculated
branch was resolved.

@@

\subsubsection{Target Function Requirements}
The benefits of \speculake rely on the computation performed in the
speculative function being well hidden. To this end we have verified 
certain requirements about the locations that a target function can 
be placed.

First and foremost the target function must be in the instruction cache. 
This does not require that the function is ever run, only loaded. We have 
found that this is not a large barrier to overcome as Intel 
processors aggressively prefetch cache pages~\cite{measuring-cache}
and simple access patterns could be established to ensure
that the required block(s) are accessible.

We have also confirmed that all speculative target code in the payload must be
stored in executable pages, and that it cannot be stored in non-executable data
sections.  We suspect that this is again due to caching. The processor has no
reason to cache instructions stored in the data section in the instruction
cache. If a trigger trains the branch predictor to jump to an address in the
payload's data section, the payload will not be able to fetch the instructions
into the execution pipeline.

>>>>

To verify the upper limit of speculative instructions, we instrumented our
trigger and payload programs to test a simple gadget of variable-length before it
communicated a signal to the real world via a cache side channel. If the cache
side channel revealed no signal in the real world, then we know the speculative
execution did not make it to the signal instructions before the mis-speculated
branch was resolved.

@@

\subsubsection{Gadget Location}
The benefits of \speculake rely on the computation performed in the
speculative function being well hidden. To this end we have verified 
certain requirements about the locations that a speculative gadget can 
be placed within a binary.

First and foremost the gadget must be in the instruction cache. 
This does not require that the actual instructions ever run,
 only that they are in cache. We have 
found that this is not a large barrier to overcome as Intel 
processors aggressively prefetch cache pages~\cite{measuring-cache}
and simple access patterns could be established to ensure
that the required block(s) are accessible.

We have also confirmed that all speculative code in the payload must be
stored in executable pages, and that it cannot be stored in non-executable data
sections. We suspect that this is again due to caching. The processor has no
reason to cache instructions stored in the data section in the instruction
cache. If a trigger trains the branch predictor to jump to an address in the
payload's data section, the payload will not be able to fetch the instructions
into the execution pipeline.

### 2018-05-08 paper/implementation.tex
target fn translated to speculative gadget

For each of these payloads, we used our custom trigger program that performs 28
indirect jumps, followed by a final indirect jump to a function pointer that
pointed to target function at address 0x430000. In each of our payload programs,
we performed the same 28 indirect jumps and the final indirect jump with the
function pointer with two differences: first, the function pointer was cleared
from the cache (using the \texttt{clflush} instruction), and second, the
function pointer pointed to a different destination (specifically, to a
function
that measured the cache side channel, recording the indexes in the probe array
that appeared to be in the cache and then flushing all indexes).
At address 0x430000, we placed our target
function code for the payload program, which is never directly called
(but executes speculatively). We note that this code must be loaded into the
cache but otherwise need not be accessed directly.

@@

We implemented a 2-symbol 5-state Busy Beaver Turing machine logic in our speculative
target function in the payload program. Our implementation reads a global state,
tape, and position, and determines how the state, tape, and position should be
updated. It encodes this update into a single byte, and loads from a global
buffer indexed by that byte, loading it into cache.
We randomized the stride length between indexes in
this array to avoid the CPU's stride prediction from preemptively loading values
in the buffer.

Our implementation is able to examine the state and current symbol
on the Turing tape and make decisions based on these values as long as they are
in cache. If one of these becomes uncached, one of two things will happen.
First, the speculative function may simply fail to report any result back to the
real world, as the uncached read takes too long to return before the speculative
world is squashed by the resolution of the function pointer. Second, the CPU may
\emph{speculate} on the value of the uncached item, which may be incorrect. Of
course, this does not affect the correctness of normal programs, but it does
impact the correctness of our speculative code. In particular, the speculative
target function will compute using an incorrect value of state or current
symbol, and ultimately compute and report the wrong update.

This error is particularly devious, as it is not an
error of bit flips or noise, but rather the processor speculating what the
speculative target function reads from memory. Thus, error correcting codes on
the reported result do not improve the situation.

@@

In order to have confidence the speculative target function executed with the
correct inputs and reported the computed result correctly, 
we repeat the execution a number of times and look for the value that was
detected as present in the cache most often over all the iterations.

@@

We discovered a candidate that corresponds to code in OpenSSL's
\texttt{nistp256.c} that contained 31 indirect jumps repeated 254 times in our
handshake. This code is used during the TLS key exchange as the server computes
the ECDHE shared secret. We made a list of 31 source-destination address pairs
for these indirect jumps, and constructed a jump/ret chain at those addresses in
our payload program. Our program jumps to the source address of a pair,
which has a single \texttt{retq} instruction, which pops the next
destination off the stack and performs the indirect jump. At each destination
address, we place a static (direct) \texttt{jmpq} to bring execution to the
next source address in our list. The final jump source/destination pair is not
performed (and its destination is not pushed to the stack). However, we place
our target function code at the destination address, which loads a specific value
into the cache.
The last source in the 31 indirect jumps performs an indirect jump that returns to our
payload program, which immediately times probing the cache to see if any value was
loaded.

>>>>

For each of these payloads, we used our custom trigger program that performs 28
indirect jumps, followed by a final indirect jump to a function pointer that
pointed to a gadget at address 0x430000. In each of our payload programs,
we performed the same 28 indirect jumps and the final indirect jump with the
function pointer with two differences: first, the function pointer was cleared
from the cache (using the \texttt{clflush} instruction), and second, the
function pointer pointed to a different destination (specifically, to a
function
that measured the cache side channel, recording the indexes in the probe array
that appeared to be in the cache and then flushing all indexes).
At address 0x430000, we placed the code for our speculative gadget within the 
payload program, which is never directly called
(but executes speculatively). We note that this code must be loaded into the
cache but otherwise need not be accessed directly.

@@

We implemented a 2-symbol 5-state Busy Beaver Turing machine logic in a  
payload application. The gadget implementation reads a global state,
tape, and position, and determines how the state, tape, and position should be
updated. It encodes this update into a single byte, and loads from a global
buffer indexed by that byte, loading it into cache.
We randomized the stride length between indexes in
this array to avoid the CPU's stride prediction from preemptively loading values
in the buffer.

Our implementation is able to examine the state and current symbol
on the Turing tape and make decisions based on these values as long as they are
in cache. If one of these becomes uncached, one of two things will happen.
First, the speculative gadget may simply fail to report any result back to the
real world, as the uncached read takes too long to return before the speculative
world is squashed by the resolution of the function pointer. Second, the CPU may
\emph{speculate} on the value of the uncached item, which may be incorrect. Of
course, this does not affect the correctness of normal programs, but it does
impact the correctness of our speculative code. In particular, the speculative
gadget will compute using an incorrect value of state or current
symbol, and ultimately report the wrong update.

This error is particularly devious, as it is not an
error of bit flips or noise, but rather the processor speculating what the
speculative gadget will read from memory. Thus, error correcting codes on
the reported result do not improve the situation.

@@

In order to have confidence the speculative gadget executed with the
correct inputs and reported the computed result correctly, 
we repeat the execution a number of times and look for the value that was
detected as present in the cache most often over all the iterations.

@@

We discovered a candidate that corresponds to code in OpenSSL's
\texttt{nistp256.c} that contained 31 indirect jumps repeated 254 times in our
handshake. This code is used during the TLS key exchange as the server computes
the ECDHE shared secret. We made a list of 31 source-destination address pairs
for these indirect jumps, and constructed a jump/ret chain at those addresses in
our payload program. Our program jumps to the source address of a pair,
which has a single \texttt{retq} instruction, which pops the next
destination off the stack and performs the indirect jump. At each destination
address, we place a static (direct) \texttt{jmpq} to bring execution to the
next source address in our list. The final jump source/destination pair is not
performed (and its destination is not pushed to the stack). However, we place
our gadget code at the destination address, which finishes by loading a specific value
into the cache.
The last source in the 31 indirect jumps performs an indirect jump that returns to our
payload program, which immediately times probing the cache to see if any value was
loaded.

### 2018-05-08 paper/introduction.tex
target fn translated to speculative gadget

Importantly, the trigger program and payload program's indirect jump patterns
diverge on the destination of their final jumps. However, because the trigger
program has trained the CPU's branch predictor, the CPU speculates that the
payload program will continue following the pattern of the trigger program,
causing it to speculatively execute a \emph{target function} in the payload
program. This execution path is shown in Figure~\ref{fig:high-level}.

We emphasize that this target function is in a region that is neither
read nor executed by the payload program, and after the CPU discovers the
mis-speculation in the payload program, it will discard the results from the
target function and continue executing from the correct destination. However,
the payload program has a limited speculative window where it can perform
computation in the target function, and can communicate results back to the
``real world'' payload program via a cache side channel.

@@

While it's possible for the trigger and payload programs to be bundled in a
single program, an analyst aware of \speculake could easily find the target
function in the payload program based on the jump pattern in the trigger program.
However, if the trigger and payload are kept separate, the analyst has a much
harder job and must identify both.
Moreover, it is possible that the trigger program be a
seemingly-unrelated benign program already on the victim's computer. We
investigate using the OpenSSL library as a benign trigger
program in Section~\ref{subsec:openssl}. If the trigger program is another benign
program on the system, the analyst has the difficult task of identifying which
program, library, or even operating system component is responsible for training
the CPU's branch predictor, and finding the specific set of jumps that occur at
runtime. To make matters worse, the payload program can include dummy jump
patterns that are effectively dead ends for the analyst, as they do not
ultimately call the target function speculatively.

>>>>

Importantly, the trigger program and payload program's indirect jump patterns
diverge on the destination of their final jumps. However, because the trigger
program has trained the CPU's branch predictor, the CPU speculates that the
payload program will continue following the pattern of the trigger program,
causing it to execute a \emph{speculative gadget} in the payload
program. This execution path is shown in Figure~\ref{fig:high-level}.

We emphasize that this gadget is in a region that is neither
read nor executed by the payload program, and after the CPU discovers the
mis-speculation in the payload program, it will discard the results from the
speculative gadget and continue executing from the correct destination. However,
the payload program has a limited speculative window where it can perform
computation, and can communicate results back to the ``real world'' payload 
program via a cache side channel.

@@

While it's possible for the trigger and payload programs to be bundled in a
single program, an analyst aware of \speculake could easily find the speculative
gadget in the payload program based on the jump pattern in the trigger program.
However, if the trigger and payload are kept separate, the analyst has a much
harder job and must identify both.
Moreover, it is possible that the trigger program be a
seemingly-unrelated benign program already on the victim's computer. We
investigate using the OpenSSL library as a benign trigger
program in Section~\ref{subsec:openssl}. If the trigger program is another benign
program on the system, the analyst has the difficult task of identifying which
program, library, or even operating system component is responsible for training
the CPU's branch predictor, and finding the specific set of jumps that occur at
runtime. To make matters worse, the payload program can include dummy jump
patterns that are effectively dead ends for the analyst, as they do not
ultimately call the hidden speculative gadget.
